Reasoning About a Machine with Local Capabilities

### 2018-04-23 toplas/paper.tex
Fix some citet's that should be citep's.

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control-flow
correctness, local-state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control-flow correctness
and local-state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) and/or rely on techniques like
machine code rewriting~\citet{wahbe_efficient_1993,abadi_control-flow_2005}, machine code
verification~\citet{morrisett_system_1999}, virtual machines with a
native stack~\citep{lindholm_java_2014} or
randomization~\citep{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

\emph{Capability machines} are a type of processors that
remediate these limitations with a better security model at
the hardware level. They are based on old ideas~\citet{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999},
but have recently received renewed interest; in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\citep{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities (a kind
of pointers that carry authority). Memory capabilities carry
the authority to read and/or write to a range of memory
locations. There is also a form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

@@

In this paper, we make two main contributions: (1) an alternative calling
convention that uses local capabilities to enforce stack frame encapsulation and
well-bracketed control flow, and (2) perhaps more importantly, we adapt and
apply the well-studied techniques of step-indexed Kripke logical relations for
reasoning about code on a representative capability machine with local
capabilities in general and correctness and security of the calling convention
in particular. More specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel step-indexed Kripke logical relation for reasoning
  about programs on the capability machine. It is an untyped logical relation,
  inspired by previous work on object capabilities \citet{Devriese:2016ObjCap}.
  We prove an analogue of the standard fundamental theorem of logical relations
  --- to the best of our knowledge, our theorem is % by far
  the most general and powerful formulation of the formal guarantees offered by
  a capability machine (a form of capability
  safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
  We are the first to apply these techniques for reasoning about capability
  machines and we believe they will prove useful for many other purposes than
  our calling convention.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem: the use of a
  \emph{single} orthogonal closure (rather than the earlier used biorthogonal
  closure) and a variant of \citet{Dreyer:jfp12}'s public and private future
  worlds~\citep{Dreyer:jfp12} to express the special nature of local
  capabilities. The logical relation and the fundamental theorem expressing
  capability safety are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

For our purposes, this restriction is unproblematic, since we are only
interested in proving semantic properties (e.g., an assertion will never fail).
However, in other situations, we may be interested in proving more syntactic
properties like the ones that are often considered in object capability
literature: confinement, no authority amplification etc. Although such
properties are more restrictive and less easy to use for reasoning,
\citet{Devriese:2016ObjCap} have demonstrated how a logical relation like
ours can be adapted to also support them, by quantifying the logical relation
over a custom interpretation of effectful computations and the type of
references. We expect their solution can be readily adapted to our setting,
modulo some details (like the fact that we do not just have read-write
capabilities, but also others).
\end{toplas}
% % \emph{No program logic} TODO: discuss this? 
\\\\
\noindent\textbf{Logical relation}\\
\emph{Single orthogonal closure} The definitions of $\stder$ and $\stdvr$ in
\figurename~\ref{fig:logrel} apply a single orthogonal closure, a new variant of an
existing pattern called biorthogonality. Biorthogonality is a pattern for
defining logical relations~\citet{krivine_classical_1994,pitts_operational_1998}
in terms of an observation relation of safe configurations (like we do). The
idea is to define safe evaluation contexts as the set of contexts that produce
safe observations when plugging safe values and define safe terms as the set of
terms that can be plugged into safe evaluation contexts to produce safe
observations. This is an alternative to more direct definitions where safe terms
are defined as terms that evaluate to safe values. An advantage of
biorthogonality is that it scales better to languages with control effects like
call/cc. Our definitions can be seen as a variant of biorthogonality, where we
take only a single orthogonal closure: we do not define safe evaluation contexts
but immediately define safe terms as those that produce safe observations when
plugged with safe values. This is natural because we model arbitrary assembly
code that does not necessarily respect a particular calling convention: return
pointers are in principle values like all others and there is no reason to treat
them specially in the logical relation.

@@

\citet{swasey:2017} have recently developed a \emph{logic}, OCPL, for verification
of object capability patterns. The logic is based on
Iris~\citep{iris,iris2,iris3}, a state of the art higher-order concurrent
separation logic and is formalized in Coq, building on the Iris Proof Mode for
Coq~\citep{ipm}. OCPL gives a more abstract and modular way of proving
capability safety for a lambda-calculus (with concurrency) compared to the
earlier work by \citet{Devriese:2016ObjCap}.
% domi: rest of the paragraph could be cut?
\begin{toplas}
In the future we would also like to
investigate a new program logic for reasoning about capability safety for our
capability machine model. We think Iris would also be a natural starting point
for such an endeavour, since Iris is really a framework, which can be
instantiated to different programming languages. OCPL was able to leverage
existing Iris specifications for the high-level language; for our capability
machine model, however, it would be necessary to devise new kinds of
specifications for our low-level programs with unstructured control-flow. It is
likely that we could get inspiration from earlier work on logics for assembly
programming languages, such as XCAP~\citet{xcap}.
\end{toplas}
\itoplas{We could further add that this is not hte level of abstraction that one would like to reason. One would like to reason about the programs we actually write - that is programs written in high-level languages. If we had a fully-abstract compiler from a nice high-level language to this low-level machine, then we could reason about the program in the high-level language and then because the compiler is fully-abstract, we would retain all the nice properties from the high-level language. How does this work fit in to this picture? The calling convention ensures two properties that we expect our nice high-level languages to have, namely well-bracketedness and local-state encapsulation. If we want to do fully-abstract compilation, then we need some way to guarantee this after compilation. Further, logical relations are often used in full-abstractness proofs. This work expands our knowledge about logical relations should be defined for low-languages and in particular capability machines. (maybe also mention that enforcement necessary for fully-abstract compilation. Could be capability, but it could also be something else - we just need something which we do not have now.)  }

>>>>

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control-flow
correctness, local-state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control-flow correctness
and local-state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) and/or rely on techniques like
machine code rewriting~\citep{wahbe_efficient_1993,abadi_control-flow_2005}, machine code
verification~\citep{morrisett_system_1999}, virtual machines with a
native stack~\citep{lindholm_java_2014} or
randomization~\citep{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

\emph{Capability machines} are a type of processors that
remediate these limitations with a better security model at
the hardware level. They are based on old ideas~\citep{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999},
but have recently received renewed interest; in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\citep{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities (a kind
of pointers that carry authority). Memory capabilities carry
the authority to read and/or write to a range of memory
locations. There is also a form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

@@

In this paper, we make two main contributions: (1) an alternative calling
convention that uses local capabilities to enforce stack frame encapsulation and
well-bracketed control flow, and (2) perhaps more importantly, we adapt and
apply the well-studied techniques of step-indexed Kripke logical relations for
reasoning about code on a representative capability machine with local
capabilities in general and correctness and security of the calling convention
in particular. More specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel step-indexed Kripke logical relation for reasoning
  about programs on the capability machine. It is an untyped logical relation,
  inspired by previous work on object capabilities \citep{Devriese:2016ObjCap}.
  We prove an analogue of the standard fundamental theorem of logical relations
  --- to the best of our knowledge, our theorem is % by far
  the most general and powerful formulation of the formal guarantees offered by
  a capability machine (a form of capability
  safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
  We are the first to apply these techniques for reasoning about capability
  machines and we believe they will prove useful for many other purposes than
  our calling convention.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem: the use of a
  \emph{single} orthogonal closure (rather than the earlier used biorthogonal
  closure) and a variant of \citet{Dreyer:jfp12}'s public and private future
  worlds~\citep{Dreyer:jfp12} to express the special nature of local
  capabilities. The logical relation and the fundamental theorem expressing
  capability safety are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

For our purposes, this restriction is unproblematic, since we are only
interested in proving semantic properties (e.g., an assertion will never fail).
However, in other situations, we may be interested in proving more syntactic
properties like the ones that are often considered in object capability
literature: confinement, no authority amplification etc. Although such
properties are more restrictive and less easy to use for reasoning,
\citet{Devriese:2016ObjCap} have demonstrated how a logical relation like
ours can be adapted to also support them, by quantifying the logical relation
over a custom interpretation of effectful computations and the type of
references. We expect their solution can be readily adapted to our setting,
modulo some details (like the fact that we do not just have read-write
capabilities, but also others).
\end{toplas}
% % \emph{No program logic} TODO: discuss this? 
\\\\
\noindent\textbf{Logical relation}\\
\emph{Single orthogonal closure} The definitions of $\stder$ and $\stdvr$ in
\figurename~\ref{fig:logrel} apply a single orthogonal closure, a new variant of an
existing pattern called biorthogonality. Biorthogonality is a pattern for
defining logical relations~\citep{krivine_classical_1994,pitts_operational_1998}
in terms of an observation relation of safe configurations (like we do). The
idea is to define safe evaluation contexts as the set of contexts that produce
safe observations when plugging safe values and define safe terms as the set of
terms that can be plugged into safe evaluation contexts to produce safe
observations. This is an alternative to more direct definitions where safe terms
are defined as terms that evaluate to safe values. An advantage of
biorthogonality is that it scales better to languages with control effects like
call/cc. Our definitions can be seen as a variant of biorthogonality, where we
take only a single orthogonal closure: we do not define safe evaluation contexts
but immediately define safe terms as those that produce safe observations when
plugged with safe values. This is natural because we model arbitrary assembly
code that does not necessarily respect a particular calling convention: return
pointers are in principle values like all others and there is no reason to treat
them specially in the logical relation.

@@

\citet{swasey:2017} have recently developed a \emph{logic}, OCPL, for verification
of object capability patterns. The logic is based on
Iris~\citep{iris,iris2,iris3}, a state of the art higher-order concurrent
separation logic and is formalized in Coq, building on the Iris Proof Mode for
Coq~\citep{ipm}. OCPL gives a more abstract and modular way of proving
capability safety for a lambda-calculus (with concurrency) compared to the
earlier work by \citet{Devriese:2016ObjCap}.
% domi: rest of the paragraph could be cut?
\begin{toplas}
In the future we would also like to
investigate a new program logic for reasoning about capability safety for our
capability machine model. We think Iris would also be a natural starting point
for such an endeavour, since Iris is really a framework, which can be
instantiated to different programming languages. OCPL was able to leverage
existing Iris specifications for the high-level language; for our capability
machine model, however, it would be necessary to devise new kinds of
specifications for our low-level programs with unstructured control-flow. It is
likely that we could get inspiration from earlier work on logics for assembly
programming languages, such as XCAP~\citep{xcap}.
\end{toplas}
\itoplas{We could further add that this is not hte level of abstraction that one would like to reason. One would like to reason about the programs we actually write - that is programs written in high-level languages. If we had a fully-abstract compiler from a nice high-level language to this low-level machine, then we could reason about the program in the high-level language and then because the compiler is fully-abstract, we would retain all the nice properties from the high-level language. How does this work fit in to this picture? The calling convention ensures two properties that we expect our nice high-level languages to have, namely well-bracketedness and local-state encapsulation. If we want to do fully-abstract compilation, then we need some way to guarantee this after compilation. Further, logical relations are often used in full-abstractness proofs. This work expands our knowledge about logical relations should be defined for low-languages and in particular capability machines. (maybe also mention that enforcement necessary for fully-abstract compilation. Could be capability, but it could also be something else - we just need something which we do not have now.)  }
